Syntax and Semantics of Cedille

### 2020-12-24 docs/semantics/paper.tex
docs: semantics: some cases for substitution

\newtheorem{theorem}{Theorem}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{observation}[theorem]{Observation}

@@

\begin{figure}
  \[
  \begin{array}{ll}
    \infer{\Gamma\vdash x\tpsynth T}{(x : t)\in\Gamma} &
    \infer{\Gamma\vdash t\tpcheck T'}{\Gamma\vdash t \tpcheck T & T' \leadsto^*_\beta T} \\ \\    
    \infer{\Gamma\vdash t\tpsynth T'}{\Gamma\vdash t \tpsynth T & T \leadsto^*_\beta T'} &
    \infer{\Gamma\vdash t\tpcheck T}{\Gamma\vdash t\tpsynth T' & T' \cong T} \\ \\    
    \infer{\Gamma\vdash \absu{\lambda}{x}{t} \tpcheck \abs{\Pi}{x}{T}{T'}}{\Gamma,x:T\vdash t\tpcheck T'} &
    \infer{\Gamma\vdash t\ t' \tpsynth [t'/x]T}{\Gamma\vdash t \tpsynth \abs{\Pi}{x}{T'}{T} & \Gamma\vdash t' \tpcheck T'} \\ \\

@@

    \infer{\Gamma\vdash [ t , t' ] \tpcheck \abs{\iota}{x}{T}{T'}}
          {\Gamma\vdash t \tpcheck T & \Gamma\vdash t' \tpcheck [t/x]T' & |t| =_{\beta\eta} |t'|} &
    \infer{\Gamma\vdash t.1 \tpsynth T}{\Gamma\vdash t \tpsynth \abs{\iota}{x}{T}{T'}} \\ \\
    \infer{\Gamma\vdash t.2 \tpsynth [t.1/x]T'}{\Gamma\vdash t \tpsynth \abs{\iota}{x}{T}{T'}} &
    \infer{\Gamma\vdash \beta\{t'\} \tpcheck \{ t_1 \simeq t_2 \}}
          {\Gamma\vdash \textit{FV}(t'\ t_1\ t_2)\subseteq \textit{dom}(\Gamma) \quad |t_1| =_{\beta\eta} |t_2|}  \\ \\    
    \infer{\Gamma\vdash \delta\ \mhyph\ t \tpcheck T}{\Gamma\vdash t\tpcheck \{ \absu{\lambda}{x}{\absu{\lambda}{y}{x}} \simeq \absu{\lambda}{x}{\absu{\lambda}{y}{y}}\}}  &
    \infer{\Gamma \vdash \rho\ t'\ @x.T\ \mhyph\ t \tpsynth [t_2/x]T}
          {
          \begin{array}{cc}
            \Gamma \vdash t' \tpsynth \{t_1 \simeq t_2\}
            & \quad \Gamma \vdash [t_2/x]T \tpsynth \star
            \\ \Gamma \vdash t \tpsynth T'
            & T' \cong [t_1/x]T
          \end{array}
          } \\ \\
    \infer{\Gamma \vdash \chi\ T\ \mhyph\ t\ \tpsynth T}
          {\Gamma \vdash T \tpsynth \star & \Gamma \vdash t \tpcheck T} &
    \infer{\Gamma\vdash \varphi\ t\ \mhyph\ t'\ \{t''\} \Leftrightarrow T}
          {\Gamma\vdash t\tpsynth \{t'\simeq t''\} & \Gamma\vdash t' \Leftrightarrow T}
  \end{array}
  \]
\caption{Rules for checking a term against a well-kinded type ($\Gamma \vdash t \tpcheck T$)
           and synthesizing a type for a term ($\Gamma \vdash t \tpsynth T$)}
\label{fig:tp}
\end{figure}

@@

Defining \verb|Top| to be $\{\absu{\lambda}{x}{x} \simeq
\absu{\lambda}{x}{x}\}$, we may assign \verb|Top| to any closed term \verb|t|,
including non-normalizing ones. In our annotated syntax, we write
\texttt{\(\beta\)\{t\}}. Even without this, the presence of $\delta$ in
combination with $\varphi$ allows us to type non-normalizing terms assuming an
erased argument $x$ of type $\{ \textit{tt} \simeq \textit{ff} \}$ for
Church-encoded booleans \textit{tt} and \textit{ff}. For example, $\delta\ \mhyph x$
has type \(\{\absu{\lambda}{x}{x} \simeq \absu{\lambda}{x}{x\ x}\}\), and with
$\varphi$ we can use this to type $\Omega$ by changing the typed term
\(\texttt{id}\ \cdot\texttt{True}\ \texttt{id}\), where \verb|True| is
\(\abs{\forall}{X}{\star}{X \to X}\). But failure of normalization does not
impinge on Theorem~\ref{thm:consis}. Extensional Martin-L\"of type theory (MLTT)
is also non-normalizing, for a very similar reason, but this fact does not contradict
its logical soundness~\cite{dybjer16}. In CDLE, the guarantees one gets about
the behavior of terms are expressed almost entirely in their types. If the types
are weak, then not much is guaranteed; but stronger types can guarantee
properties like normalization, as demonstrated by the following theorem:

@@

\startcase{.2cm}
\[
  \infer{\Gamma\vdash \beta\{t'\} \tpcheck \{ t_1 \simeq t_2 \}}
  {\Gamma\vdash \textit{FV}(t'\ t_1\ t_2)\subseteq \textit{dom}(\Gamma) \quad |t_1| =_{\beta\eta} |t_2|}
\]
Since by assumption \(|t_1| =_{\beta\eta} |t_2|\) it follows that \(\sigma|t_1|
=_{\beta\eta} \sigma|t_2|\), and also by assumption it follows that
\(\textit{FV}(t_1\ t_2) \subseteq \textit{dom}(\sigma)\).
Also, we see \(\sigma|t'|\) is a closed term.
So, $[\sigma|t'|]_{\cbe}\in\interp{\{ t_1 \simeq t_2 \}}_{\sigma,\rho}$ follows
directly from the semantics of equality types.

@@

First a few lemmas (easy proofs omitted):
\begin{lemma}
  \label{lem:inference-validity}
  \ \\
  \begin{itemize}
  \item If \(\Gamma \vdash T \tpsynth \kappa\) then \(\Gamma \vdash \kappa\)

  \item If \(\Gamma \vdash t \ T\) then \(\Gamma \vdash T \tpsynth
    \kappa\) for some \(\kappa\)
  \end{itemize}
\end{lemma}

\begin{lemma}
  \label{lem:subst-kind}
  \ \\
  \begin{itemize}
  \item If \(\Gamma,x:T \vdash T \tpsynth \kappa\) and \(\Gamma \vdash t
    \tpcheck T\) then \(\Gamma \vdash [t/x]T \tpsynth [t/x]\kappa\)

  \item If \(\Gamma,X:\kappa_2 \vdash T_1 \tpsynth \kappa_1\) and \(\Gamma
    \vdash T_2 \tpsynth \kappa_2\) then \(\Gamma \vdash [T_2/X]T_1 \tpsynth [T_2/X]\kappa_1\)
  \end{itemize}
\end{lemma}

@@

  In the other subcase, the subject of kinding is of the form
  \((\abs{\lambda}{x}{T_1}{T''})\ t\) and our assumed reduction is to
  \([t/x]T''\). By inversion of the kinding derivation, we have \(\Gamma,x:T_1
  \vdash T'' \tpsynth \kappa_1\). By Lemma~\ref{lem:subst-kind}, we have
  \(\Gamma \vdash [t/x]T'' \tpsynth [t/x]\kappa_1\).

@@

  In the other subcase, the subject of kinding is of the form
  \((\abs{\lambda}{X}{\kappa_2}{T_1''}) \cdot T_2\) and our assumed reduction is
  to \([T_2/X]T_1''\). By inversion of the kinding derivation, we have
  \(\Gamma,X:\kappa_2 \vdash T_1'' \tpsynth \kappa_1\). By
  Lemma~\ref{lem:subst-kind}, we have \(\Gamma \vdash [T_2/X]T_1'' \tpsynth
  [T_2/X]\kappa_1\).

>>>>

\newtheorem{theorem}{Theorem}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{observation}[theorem]{Observation}

@@

\begin{figure}
  \[
  \begin{array}{ll}
    \infer{\Gamma\vdash x\tpsynth T}{(x : T)\in\Gamma} &
    \infer{\Gamma\vdash t\tpcheck T'}{\Gamma\vdash t \tpcheck T & T' \leadsto^*_\beta T} \\ \\    
    \infer{\Gamma\vdash t\tpsynth T'}{\Gamma\vdash t \tpsynth T & T \leadsto^*_\beta T'} &
    \infer{\Gamma\vdash t\tpcheck T}{\Gamma\vdash t\tpsynth T' & T' \cong T} \\ \\    
    \infer{\Gamma\vdash \absu{\lambda}{x}{t} \tpcheck \abs{\Pi}{x}{T}{T'}}{\Gamma,x:T\vdash t\tpcheck T'} &
    \infer{\Gamma\vdash t\ t' \tpsynth [t'/x]T}{\Gamma\vdash t \tpsynth \abs{\Pi}{x}{T'}{T} & \Gamma\vdash t' \tpcheck T'} \\ \\

@@

    \infer{\Gamma\vdash [ t , t' ] \tpcheck \abs{\iota}{x}{T}{T'}}
          {\Gamma\vdash t \tpcheck T & \Gamma\vdash t' \tpcheck [t/x]T' & |t| =_{\beta\eta} |t'|} &
    \infer{\Gamma\vdash t.1 \tpsynth T}{\Gamma\vdash t \tpsynth \abs{\iota}{x}{T}{T'}} \\ \\
    \infer{\Gamma\vdash t.2 \tpsynth [t.1/x]T'}{\Gamma\vdash t \tpsynth \abs{\iota}{x}{T}{T'}} &
    \infer{\Gamma\vdash \beta\{t'\} \tpcheck \{ t_1 \simeq t_2 \}}
          {\textit{FV}(t')\subseteq \textit{dom}(\Gamma) \quad |t_1| =_{\beta\eta} |t_2|}  \\ \\    
    \infer{\Gamma\vdash \delta\ \mhyph\ t \tpcheck T}{\Gamma\vdash t\tpcheck \{ \absu{\lambda}{x}{\absu{\lambda}{y}{x}} \simeq \absu{\lambda}{x}{\absu{\lambda}{y}{y}}\}}  &
    \infer{\Gamma \vdash \rho\ t'\ @x.T\ \mhyph\ t \tpsynth [t_2/x]T}
          {
          \begin{array}{cc}
            \Gamma \vdash t' \tpsynth \{t_1 \simeq t_2\}
            & \quad \Gamma \vdash [t_2/x]T \tpsynth \star
            \\ \Gamma \vdash t \tpsynth T'
            & T' \cong [t_1/x]T
          \end{array}
          } \\ \\
    \infer{\Gamma \vdash \chi\ T\ \mhyph\ t\ \tpsynth T}
          {\Gamma \vdash T \tpsynth \star & \Gamma \vdash t \tpcheck T} &
    \infer{\Gamma\vdash \varphi\ t\ \mhyph\ t'\ \{t''\} \Leftrightarrow T}
          {\Gamma\vdash t\tpsynth \{t'\simeq t''\} & \Gamma\vdash t' \Leftrightarrow T}
  \end{array}
  \]
\caption{Rules for checking a term against a well-kinded type ($\Gamma \vdash t \tpcheck T$)
           and synthesizing a type for a term ($\Gamma \vdash t \tpsynth T$)}
\label{fig:tp}
\end{figure}

\begin{figure}
  \centering
  \[
    \begin{array}{lll}
      \infer{\vdash \cdot}{}
      & \infer{
        \vdash \Gamma,x:T
        }{\Gamma \vdash T \tpsynth \star}
      & \infer{
        \vdash \Gamma,X:\kappa
        }{
        \Gamma \vdash \kappa
        }
    \end{array}
  \]
  \caption{Rules for checking a context is well-formed}
  \label{fig:ctxt}
\end{figure}

@@

Defining \verb|Top| to be $\{\absu{\lambda}{x}{x} \simeq
\absu{\lambda}{x}{x}\}$, we may assign \verb|Top| to any closed term \verb|t|,
including non-normalizing ones. In our annotated syntax, we write
\texttt{\(\beta\)\{t\}}. Even without this, the presence of $\delta$ in
combination with $\varphi$ allows us to type non-normalizing terms assuming an
erased argument $x$ of type $\{ \textit{tt} \simeq \textit{ff} \}$ for
Church-encoded booleans \textit{tt} and \textit{ff}. For example, $\delta\
\mhyph\ x$ can be checked against type \(\{\absu{\lambda}{x}{x} \simeq
\absu{\lambda}{x}{x\ x}\}\), and with 
$\varphi$ we can use this to type $\Omega$ by changing the typed term
\(\texttt{id}\ \cdot\texttt{True}\ \texttt{id}\), where \verb|True| is
\(\abs{\forall}{X}{\star}{X \to X}\). But failure of normalization does not
impinge on Theorem~\ref{thm:consis}. Extensional Martin-L\"of type theory (MLTT)
is also non-normalizing, for a very similar reason, but this fact does not contradict
its logical soundness~\cite{dybjer16}. In CDLE, the guarantees one gets about
the behavior of terms are expressed almost entirely in their types. If the types
are weak, then not much is guaranteed; but stronger types can guarantee
properties like normalization, as demonstrated by the following theorem:

@@

\startcase{.2cm}
\[
  \infer{\Gamma\vdash \beta\{t'\} \tpcheck \{ t_1 \simeq t_2 \}}
  {\textit{FV}(t')\subseteq \textit{dom}(\Gamma) \quad |t_1| =_{\beta\eta} |t_2|}
\]
Since by assumption \(|t_1| =_{\beta\eta} |t_2|\) it follows that \(\sigma|t_1|
=_{\beta\eta} \sigma|t_2|\), and also by assumption it follows that
\(\textit{FV}(t_1\ t_2) \subseteq \textit{dom}(\sigma)\).
Also, we see \(\sigma|t'|\) is a closed term.
So, $[\sigma|t'|]_{\cbe}\in\interp{\{ t_1 \simeq t_2 \}}_{\sigma,\rho}$ follows
directly from the semantics of equality types.

@@

\begin{lemma}
  \label{lem:subst-class}
  Below, each statement seperately universally quantifies over meta-variables,
  and it is assumed that typing contexts occuring in assumed derivations are
  well-formed.
  We also assume types in type checking judgments are well-formed under the
  given context.
  \begin{enumerate}
  \item \textbf{Kinds:}
    \begin{itemize}
    \item If \(\Gamma_1,x:T,\Gamma_2 \vdash \kappa\) and \(\Gamma_1 \vdash t
      \tpcheck T\) then \(\Gamma_1,[t/x]\Gamma_2 \vdash [t/x] \kappa\)

    \item If \(\Gamma_1,X:\kappa',\Gamma_2 \vdash \kappa\) and \(\Gamma_1 \vdash
      T \tpsynth \kappa'\) then \(\Gamma_1,[T/X]\Gamma_2 \vdash [T/X]\kappa\)
    \end{itemize}

  \item \textbf{Types}
    \begin{itemize}
    \item If \(\Gamma_1,x:T',\Gamma_2 \vdash T \tpsynth \kappa\) and \(\Gamma_1 \vdash t
      \tpsynth T'\) then \(\Gamma_1,[t/x]\Gamma_2 \vdash [t/x]T \tpsynth [t/x]\kappa\)

    \item If \(\Gamma_1,X:\kappa_2,\Gamma_2 \vdash T_1 \tpsynth \kappa_1\) and \(\Gamma_1
      \vdash T_2 \tpsynth \kappa_2\) then \(\Gamma_1,[T_2/X]\Gamma_2 \vdash [T_2/X]T_1 \tpsynth [T_2/X]\kappa_1\)
    \end{itemize}

  \item \textbf{Terms:}
    \begin{itemize}
    \item If \(\Gamma_1,x:T',\Gamma_2 \vdash t \Leftrightarrow T\) and
      \(\Gamma_1 \vdash t' \tpsynth T'\) then \(\Gamma_1,[t'/x]\Gamma_2 \vdash [t'/x]t
      \Leftrightarrow [t'/x]T\)

    \item If \(\Gamma_1,X:\kappa,\Gamma_2 \vdash t \Leftrightarrow T'\) and
      \(\Gamma_1 \vdash T \tpsynth \kappa\) then \(\Gamma_1,[T/X]\Gamma_2 \vdash
      [T/X]t \Leftrightarrow [T/X]T'\)
    \end{itemize}
  \end{enumerate}
\end{lemma}
\begin{proof}
  By mutual induction on the assumed derivations.
  We only show a few interesting cases.

  \startcase{.2cm}
  \[
    \infer{\Gamma_1,X:\kappa,\Gamma_2 \vdash X_1 \tpsynth \kappa_1}{(X_1 : \kappa_1) \in \Gamma_1,X:\kappa,\Gamma_2}
  \]

  We have two cases.
  If \(X_1 = X\), then \(\kappa_1 = \kappa\) and by assumption \(\Gamma_1 \vdash
  T \tpsynth \kappa\), and the desired result holds by weakening.
  Otherwise, either \((X_1 : \kappa_1) \in \Gamma_1\) and \(X \notin
  \textit{FV}(\kappa_1)\), or else \((X_1 : \kappa_1) \in \Gamma_2\).
  Either way, we have \(\Gamma_1,[T/X]\Gamma_2 \vdash X_1 \tpsynth
  [T/X]\kappa_1\).

  \startcase{.2cm}
  \[ \infer{\Gamma_1,x:T,\Gamma_2\vdash T_2\ t_1 \tpsynth [t_1/x_1]\kappa
    }{
      \Gamma_1,x:T,\Gamma_2\vdash T_2 \tpsynth \abs{\Pi}{x_1}{T_1}{\kappa}
      & \Gamma_1,x:T,\Gamma_2\vdash t_1 \tpcheck T_1
    }
  \]

  By the IH, \(\Gamma_1,[t/x]\Gamma_2 \vdash [t/x]T_2 \tpsynth
  \abs{\Pi}{x_1}{[t/x]T_1}{[t/x]\kappa}\) and \(\Gamma_1,[t/x]\Gamma_2 \vdash
  [t/x]t_1 \tpcheck [t/x]T_1\).
  By the rule, we have \(\Gamma_1,[t/x]\Gamma_2 \vdash [t/x]T_2\ [t/x]t_1
  \tpsynth [[t/x]t_1/x_1][t/x]\kappa\), where the synthesized kind is equal to
  the desired \([t/x][t_1/x_1]\kappa\).

  \startcase{.2cm}
  \[
    \infer{
      \Gamma_1,x:T,\Gamma_2\vdash \{ t_1 \simeq t_2 \} : \star
    }{
      \textit{FV}(t_1\ t_2)\subseteq\textit{dom}(\Gamma_1,x:T,\Gamma_2)
    }
  \]
  It suffices to show that \(\textit{FV}([t/x]t_1\ [t/x]t_2) \subseteq
  \textit{dom}(\Gamma_1,[t/x]\Gamma_2)\).
  It is clear \(x\) is not a free variable of this expression, that the free
  variables of \(t\) are declared in \(\Gamma_1\), and by assumption the other
  free variables of it are declared in \(\Gamma_1,[t/x]\Gamma_2\).

  \startcase{.2cm}
  \[
    \infer{\Gamma_1,x:T,\Gamma_2 \vdash x_1\tpsynth T_1}{(x_1 : T_1) \in \Gamma_1,x:T,\Gamma_2}
  \]
  We elaborate on the case where \(x_1 = x\).
  It is important that we assumed that \(\Gamma_1 \vdash t \tpsynth T\) (as
  opposed to having its type checked), as we may now replace the given rule with
  this assumed derivation without changing the definition of substitution.

  \startcase{.2cm}
  \[
    \infer{\Gamma_1,x:T,\Gamma_2\vdash \beta\{t'\} \tpcheck \{ t_1 \simeq t_2 \}}
    {\textit{FV}(t')\subseteq \textit{dom}(\Gamma_1,x:T,\Gamma_2) \quad |t_1| =_{\beta\eta} |t_2|}
  \]

  From our assumptions we may conclude \(\textit{FV}([t/x]t) \subseteq
  \textit{dom}(\Gamma_1,[t/x]\Gamma_2)\), and from the second premise that
  \(|[t/x]t_1| =_{\beta\eta} |[t/x]t_2|\).
  Thus, \(\Gamma_1,[t/x]\Gamma_2\ \vdash \beta\{[t/x]t'\}\tpcheck \{[t/x]t_1
  \simeq [t/x]t_2\}\).

  \startcase{.2cm}
  \[
    \infer{\Gamma_1,x:T,\Gamma_2 \vdash \rho\ t''\ @x_1.T_1\ \mhyph\ t' \tpsynth [t_2/x_1]T_1}
    {
      \begin{array}{cc}
        \Gamma_1,x:T,\Gamma_2 \vdash t'' \tpsynth \{t_1 \simeq t_2\}
        & \quad \Gamma_1,x:T,\Gamma_2 \vdash [t_2/x_1]T_1 \tpsynth \star
        \\ \Gamma_1,x:T,\Gamma_2 \vdash t' \tpsynth T'
        & T' \cong [t_1/x_1]T_1
      \end{array}
    }
  \]
  From the IH, we have that \(\Gamma_1,[t/x]\Gamma_2 \vdash [t/x]t'' \tpsynth \{[t/x]t_1
  \simeq [t/x]t_2\}\), that \(\Gamma_1,[t/x]\Gamma_2 \vdash [t/x][t_2/x]T_1
  \tpsynth \star\), and that \(\Gamma_1,[t/x]\Gamma_2 \vdash [t/x]t' \tpsynth
  [t/x]T'\).
  From the last premise, we may conclude that \([t/x]T' \cong
  [t/x][t_1/x_1]T_1\).
  Applying the rule and permuting substitutions gives us the desired result
  (note \([t/x][t_1/x_1]T_1 = [[t/x]t_1/x_1]T_1\)).
\end{proof}

\begin{corollary}
  \ \\
  \begin{itemize}
  \item If \(\vdash \Gamma_1,x:T,\Gamma_2\) and \(\Gamma_1 \vdash t \tpsynth T\)
    then \(\vdash \Gamma_1,[t/x]\Gamma_2\)

  \item If \(\vdash \Gamma_1,X:\kappa,\Gamma_2\) and \(\Gamma_1 \vdash T
    \tpsynth \kappa\) then \(\vdash \Gamma_1,[T/X]\Gamma_2\).
  \end{itemize}
\end{corollary}
\begin{proof}
  By induction on the assumed derivation, appealing to
  Lemma~\ref{lem:subst-class} at each step.
\end{proof}

@@

  In the other subcase, the subject of kinding is of the form
  \((\abs{\lambda}{x}{T_1}{T''})\ t\) and our assumed reduction is to
  \([t/x]T''\). By inversion of the kinding derivation, we have \(\Gamma,x:T_1
  \vdash T'' \tpsynth \kappa_1\). By Lemma~\ref{lem:subst-class}, we have
  \(\Gamma \vdash [t/x]T'' \tpsynth [t/x]\kappa_1\).

@@

  In the other subcase, the subject of kinding is of the form
  \((\abs{\lambda}{X}{\kappa_2}{T_1''}) \cdot T_2\) and our assumed reduction is
  to \([T_2/X]T_1''\). By inversion of the kinding derivation, we have
  \(\Gamma,X:\kappa_2 \vdash T_1'' \tpsynth \kappa_1\). By
  Lemma~\ref{lem:subst-class}, we have \(\Gamma \vdash [T_2/X]T_1'' \tpsynth
  [T_2/X]\kappa_1\).
